Symmetry, stability, and computation of degenerate lasing modes

### 2017-01-10 referee_response.txt

We thank the referees for their helpful comments. The first referee brings up several perceived issues with our analysis, which we will address.

First, the referee questions the claim that the inversion oscillations average to zero if the beating frequencies (from the mode spacing) are much faster than the relaxation rate. This nontrivial result is the central assumption of the stationary-inversion approximation (SIA), and has been well-studied and used by multiple other authors, including in the original SALT paper (Türeci et. al. 2006 [ref. 8 in our manuscript]). There, the authors cite (Fu and Haken 1991 [ref. 26 in our manuscript]), who were the first to argue that the SIA is valid for multimode lasers. An analytical proof can also be found in (Haken 1985 [ref. 24 in our manuscript]) in Eq. 6.76 and Eq. 6.77 of that work, after which Haken directly states our claim: "the pulsations are negligible if the frequency distance of the laser modes is large compared to the inverse of the longitudinal relaxation time T of the inversion." A thorough quantitative verification of SALT and the SIA in the multimode regime was performed in (Ge et. al. 2008 [ref. 11 in our manuscript]) using full time-domain simulations of the original Maxwell--Bloch equations. We hope all of these references are sufficient to convince the reader that the SIA (which hinges on, and is equivalent to having the inversion oscillations average to zero) has been well-studied, and its regime of validity well-documented. To make it clearer that our assumption of SIA is nontrivial and based on previous work, we have slightly rearranged the placement of these references around the sentence in question in our manuscript, as well as added a short description of the result from (Haken 1985) that makes it clearer why having far-apart modes results in a stationary inversion.

The referee also mentions the need to separate the inversion into a slowly-varying term and terms oscillating at the beating frequencies. This is valid and can be done, but is not necessary in our case since it would simply result in an approximately stationary inversion. The point is that in the regime where the mode spacing is much larger than the inversion relaxation rate, the "slowly-varying term" would consequently not vary at all and stay fixed at a non-zero value, while the oscillating terms would still oscillate, but with such a small amplitude as to not have a noticeable effect on the overall inversion. In fact, Eq. 6.76 in (Haken 1985) specifically says that the ratio of the amplitude of these beating oscillations to the steady-state term is very small, provided that the beating is much faster than the relaxation rate.

Finally, the referee brings up the point that our analysis does not apply when two modes are closely-spaced but not exactly degenerate far above threshold. We do not dispute this: in fact, not only our analysis, but the entire SALT formalism would break down if this were to happen. This is why we have stated in multiple places in our manuscript the regime under which we are working (well-separated modes), which also is in fact an assumption that is made when non-degenerate SALT is used, and which can if necessary be checked for consistency, just as in non-degenerate SALT. For completeness, we now address the two ways that one might expect there to be closely-spaced modes far above threshold in the case of Cnv geometry. First, that a circulating lasing mode may split into two, for some reason. This is explicitly ruled out by the fact that the degeneracy persists exactly, without splitting, even above threshold (no matter how high, as long as the intensity pattern preserves Cn symmetry), which we have explained extensively in Sec. IV and Appendix A. Second, the possibility that a second degenerate pair may start lasing at some frequency far away from that of the first pair, but then converge towards the first pair as the pump strength is increased. We have found empirically that lasing frequencies even far above threshold don't change appreciably enough for this to happen. Additionally, even if it did, it would simply result in SALT breaking down (a regime that we have clearly stated we are not working in), and that possibility can easily be accounted for by monitoring the spacing between lasing frequencies just as in non-degenerate SALT; the presence of degeneracies does not require any extra precautions to be taken. For clarity, we have added short reminder in Sec. IV A. stating that when multiple pairs of modes are lasing, one technically needs to monitor the mode spacing to make sure stationary inversion is satisfied, just as in the case of non-degenerate lasers.

We hope that we have satisfactorily addressed the referee's concerns with our explanations and revisions, and as always, we welcome any further comments and suggestions.

### 2017-02-07 degeneracy_paper.tex
finished last edits

\selectlanguage{english}%

@@

\omega & =\omega_{\mathrm{t}}+\omega_{1}d+\mathcal{O}(d^{2})\label{eq:lasing_ansatz}

@@

\mathsf{A} & =\mathsf{A}_{0}+\mathsf{A}_{1/2}\sqrt{d}+\mathsf{A}_{1}d+\mathcal{O}(d^{3/2})\nonumber \\
\mathsf{B} & =\mathsf{B}_{0}+\mathsf{B}_{1}d+\mathcal{O}(d^{2})\label{eq:matrix expansion}\\

@@

\mathsf{x} & =\mathsf{x}_{0}+\mathsf{x}_{1/2}\sqrt{d}+\mathsf{x}_{1}d+\mathcal{O}(d^{3/2})\nonumber \\
\sigma & =\sigma_{0}+\sigma_{1/2}\sqrt{d}+\sigma_{1}d+\mathcal{O}(d^{3/2}).\label{eq:eigenpair expansion}

@@

\begin{figure}
\centerline{\includegraphics[scale=0.27]{redeps}$\quad$\includegraphics[scale=0.27]{imdeps}}\caption{(Color online) Dielectric perturbation $\delta\varepsilon$ obtained
by solving QP for threshold modes with $\ell=8$. The real part (left)
has a dependence $\cos(2\ell\phi)$, while the imaginary part (right)
is a more complicated function. \label{fig:even_cylinder_deps}}
\end{figure}
\begin{figure}
\centerline{\includegraphics[bb=0bp 160bp 612bp 630bp,clip,scale=0.4]{qp}}

@@

An example of a hexagonal cavity is shown in Fig.~\ref{fig:overview}.
This geometry was adapted from an infinite lattice of period $a$
with air holes of radius $0.3a$. A single hole in the middle has
a reduced radius $0.2a$ to create a defect in the band gap. The dielectric
is $\varepsilon_{c}=11.56$ everywhere except in the holes, where
there is air. A perfectly matched layer (PML) is added to the boundaries
to simulate the radiation loss, and the axes of the hexagon have been
aligned with the diagonals rather than the $x$ and $y$ axes because
the finite-difference Yee discretization \cite{taflove} happens to
only have mirror symmetry along the diagonals. Here, the lasing modes
are TE (electric field in-plane and magnetic field out of plane),
and there is a pair of degenerate threshold modes from the hexagon's
$C_{6\mathrm{v}}$ symmetry, as shown in Fig.~\ref{fig:overview}.
For a $100\times100$ finite-difference discretization, there is about
a 1.5\% splitting between the threshold eigenvalues, so again we must
use the QP procedure to force the threshold degeneracy. Since these
are TE modes, we now have \emph{two }components of the electric field,
and consequently we may treat $\delta\varepsilon$ as a tensor, as
in Eq.~(\ref{eq:tensor_delta_epsilon}). We only consider the diagonal
components $\delta\varepsilon_{xx}$ and $\delta\varepsilon_{yy}$
here for simplicity. Only two iterations of QP are necessary to force
the degeneracy down to machine precision, and the perturbation used
to force the degeneracy is shown in Fig.~\ref{fig:phc_deps}. We
then use Eq.~(\ref{eq:chiral definition}) as an initial guess for
our numerical solver, and the intensity pattern of the resulting circulating
solution is shown in Fig.~\ref{fig:phc_stable}. 
\begin{figure}
\centerline{\includegraphics[scale=0.27]{phc_depsxxR}$\qquad$\includegraphics[scale=0.27]{phc_depsxxI}}\caption{(Color online) Dielectric perturbation obtained from QP procedure
for hexagonal cavity. Since the mode is TE ($\mathbf{E}=E_{\mathrm{even}}\hat{\mathbf{x}}+E_{\mathrm{odd}}\hat{\mathbf{y}}$),
we have allowed the perturbation to be a diagonally-anisotropic tensor,
as in Eq.~(\ref{eq:tensor_delta_epsilon}). Shown here are the real
(left) and imaginary (right) parts of $\delta\varepsilon_{xx}$. The
$\delta\varepsilon_{yy}$ looks similar except rotated by 60 degrees.
\label{fig:phc_deps}}
\end{figure}
\begin{figure}
\centerline{\includegraphics[scale=0.31]{phc_stableIntensity}}\caption{(Color online) Intensity pattern for stable lasing mode for hexagonal
cavity. The pattern appears to be six-fold symmetric, which is expected.
Unlike in the right panel of Fig.~\ref{fig:lowQ_chirality} however,
the chirality is not significant enough to be visible, since the hexagonal
cavity is not as lossy as the square cavity in Fig.~\ref{fig:lowQ_chirality}.
In the ideal system, the second pole $\delta\omega^{\prime}$ stays
degenerate with the lasing eigenvalue $\delta\omega$, and this linear
combination stays stable for all pump strengths above threshold. In
the discretized system, there is not a true $C_{6\mathrm{v}}$ symmetry,
so there is a small splitting similar to that of the even-$\ell$
cylinder modes. Again, this splitting is too small to affect physically
meaningful results of the simulation, but can be removed using the
QP procedure if desired. \label{fig:phc_stable}}
\end{figure}

@@

that $\mathbf{E}=\mathcal{O}(\sqrt{d})$, and that the profile should
be some linear combination of the threshold modes $\mathbf{E}_{1,2}$,
as in Eq.~(\ref{eq:lasing_ansatz}). Inserting Eq.~(\ref{eq:lasing_ansatz})

@@

\begin{equation}
\mathsf{A}=\left(\begin{array}{ccccc}
\Delta & -\varepsilon_{\mathrm{I}}\omega^{2} & \omega^{2} & 0 & 0\\
\varepsilon_{\mathrm{I}}\omega^{2} & \Delta & 0 & \omega^{2} & 0\\
\gamma_{\perp}D & 0 & \omega_{a}-\omega & \gamma_{\perp} & \gamma_{\perp}\mathbf{E}_{\mathrm{R}}\\
0 & \gamma_{\perp}D & -\gamma_{\perp} & \omega_{a}-\omega & \gamma_{\perp}\mathbf{E}_{\mathrm{I}}\\
-\mathbf{P}_{\mathrm{I}} & \mathbf{P}_{\mathrm{R}} & \mathbf{E}_{\mathrm{I}} & -\mathbf{E}_{\mathrm{R}} & \gamma_{\parallel}
\end{array}\right),\label{eq:A matrix}
\end{equation}
where $\Delta=\varepsilon_{\mathrm{R}}\omega^{2}-\nabla\times\nabla\times$,
$\mathbf{E}_{\mathrm{R}}\equiv\Re(\mathbf{E})$, and $\mathbf{E}_{\mathrm{I}}\equiv\Im(\mathbf{E})$,
\begin{equation}
\mathsf{B}=\left(\begin{array}{ccccc}
-2\varepsilon_{\mathrm{I}}\omega & -2\varepsilon_{\mathrm{R}}\omega & 0 & -2\omega & 0\\
2\varepsilon_{\mathrm{R}}\omega & -2\varepsilon_{\mathrm{I}}\omega & 2\omega & 0 & 0\\
0 & 0 & 0 & 1 & 0\\
0 & 0 & -1 & 0 & 0\\
0 & 0 & 0 & 0 & 1
\end{array}\right),
\end{equation}
and
\begin{equation}
\mathsf{C}=\left(\begin{array}{ccccc}
-\varepsilon_{\mathrm{R}} & \varepsilon_{\mathrm{I}} & -1 & 0 & 0\\
-\varepsilon_{\mathrm{I}} & -\varepsilon_{\mathrm{R}} & 0 & -1 & 0\\
0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0
\end{array}\right).
\end{equation}
We expand the matrices in powers of $\sqrt{d}$, as in Eq.~(\ref{eq:matrix expansion}),
by noting that $\omega=\omega_{\mathrm{t}}+\omega_{1}d+\mathcal{O}(d^{2})$,
$D_{0}=D_{\mathrm{t}}(1+d)$, and $\mathbf{E},\mathbf{P}=\mathcal{O}(\sqrt{d})$,
according to Eq.~(\ref{eq:lasing_ansatz}). Now expand Eq.~(\ref{eq:quadratic eigenproblem})
and keep track of terms order-by-order using Eq.~(\ref{eq:eigenpair expansion}).

@@

At the next order, $\mathcal{O}(d)$, Eq.~(\ref{eq:quadratic eigenproblem})
is
\begin{equation}
(\mathsf{B}_{0}\sigma_{1}+\mathsf{A}_{1})\sum b_{k}\mathsf{v}_{k}+\mathsf{A}_{1/2}\mathsf{x}_{1/2}+\mathsf{A}_{0}\mathsf{x}_{1}=0.\label{eq:first order quadratic}
\end{equation}
(Note that without the $\mathsf{A}_{1/2}\mathsf{x}_{1/2}$ term, Eq.~(\ref{eq:first order quadratic})
would yield identical results to finding the two passive poles of
the SALT equation that come from the threshold degenerate lasing modes.)
Again, act on this equation with $\mathsf{w}_{j}^{T}$. By direct
evaluation, we have
\begin{multline}
\mathsf{w}_{j}^{T}\mathsf{A}_{1}\mathsf{v}_{k}=\Re\left[\omega_{1}\frac{\partial}{\partial\omega_{\mathrm{t}}}\omega_{\mathrm{t}}^{2}\int d^{3}x\,\vec{e}_{j}\cdot(\varepsilon+\Gamma_{\mathrm{t}}D_{\mathrm{t}})\vec{e}_{k}\right]+\\
\Re\left[\omega_{\mathrm{t}}^{2}\Gamma_{\mathrm{t}}\int d^{3}x\,D_{\mathrm{t}}\vec{e}_{j}\cdot\left(1-\left|\vec{f}\right|^{2}\right)\vec{e}_{k}\right]
\end{multline}
and
\begin{equation}
\mathsf{w}_{j}^{T}\mathsf{B}_{0}\mathsf{v}_{k}=-\Im\left[\frac{\partial}{\partial\omega_{\mathrm{t}}}\omega_{\mathrm{t}}^{2}\int d^{3}x\,\vec{e}_{j}\cdot(\varepsilon+\Gamma_{\mathrm{t}}D_{\mathrm{t}})\vec{e}_{k}\right].
\end{equation}
 By the same symmetry arguments used to evaluate the integrals in
Appendix~\ref{sec: appendix allowed lasing modes}, it is straightforward
to show that 
\begin{equation}
\frac{\partial}{\partial\omega_{\mathrm{t}}}\omega_{\mathrm{t}}^{2}\int d^{3}x\,\vec{e}_{j}\cdot(\varepsilon+\Gamma_{\mathrm{t}}D_{\mathrm{t}})\vec{e}_{k}=\omega_{\mathrm{t}}^{2}\Gamma_{\mathrm{t}}H\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk},
\end{equation}
where $H$ is given in Eq.~(\ref{eq:G and H definitions}) and
\begin{equation}
\mathsf{X}\equiv\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right).
\end{equation}
Next, we have
\begin{multline}
\int d^{3}x\,D_{\mathrm{t}}\vec{e}_{j}\cdot\left|\vec{f}\right|^{2}\vec{e}_{k}=\left(\begin{array}{cc}
\mathsf{M} & i\mathsf{M}\\
i\mathsf{M} & -\mathsf{M}
\end{array}\right)_{jk}+\\
\left(\left|a_{+}\right|^{2}I_{+}+\left|a_{-}\right|^{2}I_{-}\right)\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk}
\end{multline}
where
\begin{equation}
\mathsf{M}\equiv\left(\begin{array}{cc}
a_{+}^{\star}a_{-}J_{+}+a_{+}a_{-}^{\star}K_{+} & 0\\
0 & a_{+}^{\star}a_{-}K_{-}+a_{+}a_{-}^{\star}J_{-}
\end{array}\right).
\end{equation}
Putting these results together, we have
\begin{multline}
\mathsf{w}_{j}^{T}\mathsf{A}_{1}\mathsf{v}_{k}=\\
\omega_{\mathrm{t}}^{2}\Re\left[\Gamma_{\mathrm{t}}W\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk}-\Gamma_{\mathrm{t}}\left(\begin{array}{cc}
\mathsf{M} & i\mathsf{M}\\
i\mathsf{M} & -\mathsf{M}
\end{array}\right)_{jk}\right]\label{eq:A1 element final}
\end{multline}
where 
\begin{equation}
W\equiv\omega_{\mathrm{1}}H+G_{D}-\left|a_{+}\right|^{2}I_{+}-\left|a_{-}\right|^{2}I_{-},
\end{equation}
and
\begin{equation}
\mathsf{w}_{j}^{T}\mathsf{B}_{0}\mathsf{v}_{k}=-\omega_{\mathrm{t}}^{2}\Im\left[\Gamma_{\mathrm{t}}H\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk}\right].\label{eq:B0 element final}
\end{equation}

@@

and $\mathsf{f}_{3/2}$ is $\mathcal{O}(1)$ with respect to $\gamma_{\parallel}$,

@@

of $\mathcal{O}(\gamma_{\parallel}^{-1})$: $-\mathsf{A}_{1/2}\mathsf{x}_{3/2}=\gamma_{\parallel}^{-1}\mathsf{A}_{1/2}\mathsf{f}_{3/2}$.

@@

\gamma_{\parallel}\sigma_{2}=\frac{\mathsf{x}_{0}^{T}\mathsf{A}_{1/2}\mathsf{f}_{3/2}}{\mathsf{x}_{0}^{T}\mathsf{B}_{0}\mathsf{x}_{0}}+\mathcal{O}(\gamma_{\parallel}).

@@

where now $\mathsf{f}_{5/2}=\mathcal{O}(\gamma_{\parallel}^{-1})$
due to it including terms with $\sigma_{2}$ and $\mathsf{x}_{2}$.
By continuing this process, we find that $\sigma_{m+\frac{1}{2}}=0$,
and $\mathsf{x}_{m+\frac{1}{2}}$, $\sigma_{m+1}$, and $\mathsf{x}_{m+1}$
are all $\mathcal{O}(\gamma_{\parallel}^{-m})$. This is made possible
by the fact that the only place $\gamma_{\parallel}$ occurs in the
entire problem is the very last matrix element of $\mathsf{A}_{0}$

@@

from those of all the other matrices. Defining $s_{m+1}\equiv\gamma_{\parallel}^{m}\sigma_{m+1}=\mathcal{O}(1)$,

>>>>

@@

\omega & =\omega_{\mathrm{t}}+\omega_{1}d+O(d^{2})\label{eq:lasing_ansatz}

@@

\mathsf{A} & =\mathsf{A}_{0}+\mathsf{A}_{1/2}\sqrt{d}+\mathsf{A}_{1}d+O(d^{3/2})\nonumber \\
\mathsf{B} & =\mathsf{B}_{0}+\mathsf{B}_{1}d+O(d^{2})\label{eq:matrix expansion}\\

@@

\mathsf{x} & =\mathsf{x}_{0}+\mathsf{x}_{1/2}\sqrt{d}+\mathsf{x}_{1}d+O(d^{3/2})\nonumber \\
\sigma & =\sigma_{0}+\sigma_{1/2}\sqrt{d}+\sigma_{1}d+O(d^{3/2}).\label{eq:eigenpair expansion}

@@

\begin{figure}
\centerline{\includegraphics[scale=0.1]{redeps}$\quad$\includegraphics[scale=0.1]{imdeps}}\caption{(Color online) Dielectric perturbation $\delta\varepsilon$ obtained
by solving QP for threshold modes with $\ell=8$. The real part (left)
has a dependence $\cos(2\ell\phi)$, while the imaginary part (right)
is a more complicated function. \label{fig:even_cylinder_deps}}
\end{figure}
\begin{figure}
\centerline{\includegraphics[bb=0bp 160bp 612bp 630bp,clip,scale=0.4]{qp}}

@@

An example of a hexagonal cavity is shown in Fig.~\ref{fig:overview}.
This geometry was adapted from an infinite lattice of period $a$
with air holes of radius $0.3a$. A single hole in the middle has
a reduced radius $0.2a$ to create a defect in the band gap. The dielectric
is $\varepsilon_{c}=11.56$ everywhere except in the holes, where
there is air. A perfectly matched layer (PML) is added to the boundaries
to simulate the radiation loss, and the axes of the hexagon have been
aligned with the diagonals rather than the $x$ and $y$ axes because
the finite-difference Yee discretization \cite{taflove} happens to
only have mirror symmetry along the diagonals. Here, the lasing modes
are TE (electric field in-plane and magnetic field out of plane),
and there is a pair of degenerate threshold modes from the hexagon's
$C_{6\mathrm{v}}$ symmetry, as shown in Fig.~\ref{fig:overview}.
For a $100\times100$ finite-difference discretization, there is about
a 1.5\% splitting between the threshold eigenvalues, so again we must
use the QP procedure to force the threshold degeneracy. Since these
are TE modes, we now have \emph{two }components of the electric field,
and consequently we may treat $\delta\varepsilon$ as a tensor, as
in Eq.~(\ref{eq:tensor_delta_epsilon}). We only consider the diagonal
components $\delta\varepsilon_{xx}$ and $\delta\varepsilon_{yy}$
here for simplicity. Only two iterations of QP are necessary to force
the degeneracy down to machine precision, and the perturbation used
to force the degeneracy is shown in Fig.~\ref{fig:phc_deps}. We
then use Eq.~(\ref{eq:chiral definition}) as an initial guess for
our numerical solver, and the intensity pattern of the resulting circulating
solution is shown in Fig.~\ref{fig:phc_stable}. 
\begin{figure}
\centerline{\includegraphics[scale=0.1]{phc_depsxxR}$\qquad$\includegraphics[scale=0.1]{phc_depsxxI}}\caption{(Color online) Dielectric perturbation obtained from QP procedure
for hexagonal cavity. Since the mode is TE ($\mathbf{E}=E_{\mathrm{even}}\hat{\mathbf{x}}+E_{\mathrm{odd}}\hat{\mathbf{y}}$),
we have allowed the perturbation to be a diagonally-anisotropic tensor,
as in Eq.~(\ref{eq:tensor_delta_epsilon}). Shown here are the real
(left) and imaginary (right) parts of $\delta\varepsilon_{xx}$. The
$\delta\varepsilon_{yy}$ looks similar except rotated by 60 degrees.
\label{fig:phc_deps}}
\end{figure}
\begin{figure}
\centerline{\includegraphics[scale=0.31]{phc_stableIntensity}}\caption{(Color online) Intensity pattern for stable lasing mode for hexagonal
cavity. The pattern appears to be six-fold symmetric, which is expected.
Unlike in the right panel of Fig.~\ref{fig:lowQ_chirality} however,
the chirality is not significant enough to be visible, since the hexagonal
cavity is not as lossy as the square cavity in Fig.~\ref{fig:lowQ_chirality}.
In the ideal system, the second pole $\delta\omega^{\prime}$ stays
degenerate with the lasing eigenvalue $\delta\omega$, and this linear
combination stays stable for all pump strengths above threshold. In
the discretized system, there is not a true $C_{6\mathrm{v}}$ symmetry,
so there is a small splitting similar to that of the even-$\ell$
cylinder modes. Again, this splitting is too small to affect physically
meaningful results of the simulation, but can be removed using the
QP procedure if desired. \label{fig:phc_stable}}
\end{figure}

@@

that $\mathbf{E}=O(\sqrt{d})$, and that the profile should be some
linear combination of the threshold modes $\mathbf{E}_{1,2}$, as
in Eq.~(\ref{eq:lasing_ansatz}). Inserting Eq.~(\ref{eq:lasing_ansatz})

@@

\begin{equation}
\mathsf{A}=\left(\begin{array}{ccccc}
\Delta & -\varepsilon_{\mathrm{I}}\omega^{2} & \omega^{2} & 0 & 0\\
\varepsilon_{\mathrm{I}}\omega^{2} & \Delta & 0 & \omega^{2} & 0\\
\gamma_{\perp}D & 0 & \omega_{a}-\omega & \gamma_{\perp} & \gamma_{\perp}\mathbf{E}_{\mathrm{R}}\\
0 & \gamma_{\perp}D & -\gamma_{\perp} & \omega_{a}-\omega & \gamma_{\perp}\mathbf{E}_{\mathrm{I}}\\
-\mathbf{P}_{\mathrm{I}} & \mathbf{P}_{\mathrm{R}} & \mathbf{E}_{\mathrm{I}} & -\mathbf{E}_{\mathrm{R}} & \gamma_{\parallel}
\end{array}\right),\label{eq:A matrix}
\end{equation}
where $\Delta=\varepsilon_{\mathrm{R}}\omega^{2}-\nabla\times\nabla\times$,
$\mathbf{E}_{\mathrm{R}}\equiv\Re(\mathbf{E})$, and $\mathbf{E}_{\mathrm{I}}\equiv\Im(\mathbf{E})$,
\begin{equation}
\mathsf{B}=\left(\begin{array}{ccccc}
-2\varepsilon_{\mathrm{I}}\omega & -2\varepsilon_{\mathrm{R}}\omega & 0 & -2\omega & 0\\
2\varepsilon_{\mathrm{R}}\omega & -2\varepsilon_{\mathrm{I}}\omega & 2\omega & 0 & 0\\
0 & 0 & 0 & 1 & 0\\
0 & 0 & -1 & 0 & 0\\
0 & 0 & 0 & 0 & 1
\end{array}\right),
\end{equation}
and
\begin{equation}
\mathsf{C}=\left(\begin{array}{ccccc}
-\varepsilon_{\mathrm{R}} & \varepsilon_{\mathrm{I}} & -1 & 0 & 0\\
-\varepsilon_{\mathrm{I}} & -\varepsilon_{\mathrm{R}} & 0 & -1 & 0\\
0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 0
\end{array}\right).
\end{equation}
We expand the matrices in powers of $\sqrt{d}$, as in Eq.~(\ref{eq:matrix expansion}),
by noting that $\omega=\omega_{\mathrm{t}}+\omega_{1}d+O(d^{2})$,
$D_{0}=D_{\mathrm{t}}(1+d)$, and $\mathbf{E},\mathbf{P}=O(\sqrt{d})$,
according to Eq.~(\ref{eq:lasing_ansatz}). Now expand Eq.~(\ref{eq:quadratic eigenproblem})
and keep track of terms order-by-order using Eq.~(\ref{eq:eigenpair expansion}).

@@

At the next order, $O(d)$, Eq.~(\ref{eq:quadratic eigenproblem})
is
\begin{equation}
(\mathsf{B}_{0}\sigma_{1}+\mathsf{A}_{1})\sum b_{k}\mathsf{v}_{k}+\mathsf{A}_{1/2}\mathsf{x}_{1/2}+\mathsf{A}_{0}\mathsf{x}_{1}=0.\label{eq:first order quadratic}
\end{equation}
(Note that without the $\mathsf{A}_{1/2}\mathsf{x}_{1/2}$ term, Eq.~(\ref{eq:first order quadratic})
would yield identical results to finding the two passive poles of
the SALT equation that come from the threshold degenerate lasing modes.)
Again, act on this equation with $\mathsf{w}_{j}^{T}$. By direct
evaluation, we have
\begin{multline}
\mathsf{w}_{j}^{T}\mathsf{A}_{1}\mathsf{v}_{k}=\Re\left[\omega_{1}\frac{\partial}{\partial\omega_{\mathrm{t}}}\omega_{\mathrm{t}}^{2}\int d^{3}x\,\vec{e}_{j}\cdot(\varepsilon+\Gamma_{\mathrm{t}}D_{\mathrm{t}})\vec{e}_{k}\right]+\\
\Re\left[\omega_{\mathrm{t}}^{2}\Gamma_{\mathrm{t}}\int d^{3}x\,D_{\mathrm{t}}\vec{e}_{j}\cdot\left(1-\left|\vec{f}\right|^{2}\right)\vec{e}_{k}\right]
\end{multline}
and
\begin{equation}
\mathsf{w}_{j}^{T}\mathsf{B}_{0}\mathsf{v}_{k}=-\Im\left[\frac{\partial}{\partial\omega_{\mathrm{t}}}\omega_{\mathrm{t}}^{2}\int d^{3}x\,\vec{e}_{j}\cdot(\varepsilon+\Gamma_{\mathrm{t}}D_{\mathrm{t}})\vec{e}_{k}\right].
\end{equation}
 By the same symmetry arguments used to evaluate the integrals in
Appendix~\ref{sec: appendix allowed lasing modes}, it is straightforward
to show that 
\begin{equation}
\frac{\partial}{\partial\omega_{\mathrm{t}}}\omega_{\mathrm{t}}^{2}\int d^{3}x\,\vec{e}_{j}\cdot(\varepsilon+\Gamma_{\mathrm{t}}D_{\mathrm{t}})\vec{e}_{k}=\omega_{\mathrm{t}}^{2}\Gamma_{\mathrm{t}}H\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk},
\end{equation}
where $H$ is given in Eq.~(\ref{eq:G and H definitions}) and
\begin{equation}
\mathsf{X}\equiv\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right).
\end{equation}
Next, we have
\begin{multline}
\int d^{3}x\,D_{\mathrm{t}}\vec{e}_{j}\cdot\left|\vec{f}\right|^{2}\vec{e}_{k}=\left(\begin{array}{cc}
\mathsf{M} & i\mathsf{M}\\
i\mathsf{M} & -\mathsf{M}
\end{array}\right)_{jk}+\\
\left(\left|a_{+}\right|^{2}I_{+}+\left|a_{-}\right|^{2}I_{-}\right)\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk}
\end{multline}
where
\begin{equation}
\mathsf{M}\equiv\left(\begin{array}{cc}
a_{+}^{\star}a_{-}J_{+}+a_{+}a_{-}^{\star}K_{+} & 0\\
0 & a_{+}^{\star}a_{-}K_{-}+a_{+}a_{-}^{\star}J_{-}
\end{array}\right).
\end{equation}
Putting these results together, we have
\begin{multline}
\mathsf{w}_{j}^{T}\mathsf{A}_{1}\mathsf{v}_{k}=\\
\omega_{\mathrm{t}}^{2}\Re\left[\Gamma_{\mathrm{t}}W\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk}-\Gamma_{\mathrm{t}}\left(\begin{array}{cc}
\mathsf{M} & i\mathsf{M}\\
i\mathsf{M} & -\mathsf{M}
\end{array}\right)_{jk}\right]\label{eq:A1 element final}
\end{multline}
where 
\begin{equation}
W\equiv\omega_{\mathrm{1}}H+G_{D}-\left|a_{+}\right|^{2}I_{+}-\left|a_{-}\right|^{2}I_{-},
\end{equation}
and
\begin{equation}
\mathsf{w}_{j}^{T}\mathsf{B}_{0}\mathsf{v}_{k}=-\omega_{\mathrm{t}}^{2}\Im\left[\Gamma_{\mathrm{t}}H\left(\begin{array}{cc}
\mathsf{X} & i\mathsf{X}\\
i\mathsf{X} & -\mathsf{X}
\end{array}\right)_{jk}\right].\label{eq:B0 element final}
\end{equation}

@@

and $\mathsf{f}_{3/2}$ is $O(1)$ with respect to $\gamma_{\parallel}$,

@@

of $O(\gamma_{\parallel}^{-1})$: $-\mathsf{A}_{1/2}\mathsf{x}_{3/2}=\gamma_{\parallel}^{-1}\mathsf{A}_{1/2}\mathsf{f}_{3/2}$.

@@

\gamma_{\parallel}\sigma_{2}=\frac{\mathsf{x}_{0}^{T}\mathsf{A}_{1/2}\mathsf{f}_{3/2}}{\mathsf{x}_{0}^{T}\mathsf{B}_{0}\mathsf{x}_{0}}+O(\gamma_{\parallel}).

@@

where now $\mathsf{f}_{5/2}=O(\gamma_{\parallel}^{-1})$ due to it
including terms with $\sigma_{2}$ and $\mathsf{x}_{2}$. By continuing
this process, we find that $\sigma_{m+\frac{1}{2}}=0$, and $\mathsf{x}_{m+\frac{1}{2}}$,
$\sigma_{m+1}$, and $\mathsf{x}_{m+1}$ are all $O(\gamma_{\parallel}^{-m})$.
This is made possible by the fact that the only place $\gamma_{\parallel}$
occurs in the entire problem is the very last matrix element of $\mathsf{A}_{0}$

@@

from those of all the other matrices. Defining $s_{m+1}\equiv\gamma_{\parallel}^{m}\sigma_{m+1}=O(1)$,
